### 2014-08-20 doc/thesis/system.tex
Reorganizing the sections of Chapter 5 to start writing it
Cosmetic changes to chapters 4 and 5

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; P \rightarrow R & \textit{function types}\\
& | \; \{ \}_{o|c} & \textit{empty table type}\\
& | \; \{F, ..., F\}_{o|c} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_{i} & \textit{projection types}\\
L ::= & \False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} & \\
B ::= & \Boolean \; | \; \Number \; | \; \String & \\
F ::= & K:T \; | \; \Const \; K:T & \textit{field types}\\ 
K ::= & L \; | \; B \; | \; \Value \; | \; \Any & \textit{key types}\\
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
S ::= & \;\; T &\\
& | \; T* & \textit{vararg types}\\
& | \; T \times S & \textit{tuple types}\\
P ::= & \Void \; | \; S & \textit{parameters list type}\\
R ::= & \Void \; | \; S \; | \; S \cup R & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

@@

\senv \vdash {<}{\it number}{>} \subtype \Number

@@

\senv \vdash {<}{\it string}{>} \subtype \String

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = \vec{e}  \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; \vec{e} \; | \;
a\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{<}{\it number}{>} \; | \;
{<}{\it string}{>} \; | \;
{...} \; | \;
f \; | \;
\{ \; \vec{c}\; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
a \; | \;
l \\
l ::= & \; n \; | \;
e_{1}[e_{2}]\\
a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
e_{1}{:}n(\vec{e_{2}})\\
f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
c ::= & \; [e_{1}] = e_{2}\\
n ::= & \; {<}{\it name}{>}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

\senv \vdash {<}{\it number}{>} \lesssim \Number

@@

\senv \vdash {<}{\it string}{>} \lesssim \String

@@

>>>>

\section{Type Language}

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; P \rightarrow R & \textit{function types}\\
& | \; \{ \}_{o|c} & \textit{empty table type}\\
& | \; \{F, ..., F\}_{o|c} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_{i} & \textit{projection types}\\
L ::= & \False \; | \; \True \; | \; {\it number} \; | \; {\it string} & \\
B ::= & \Boolean \; | \; \Number \; | \; \String & \\
F ::= & K:T \; | \; \Const \; K:T & \textit{field types}\\ 
K ::= & L \; | \; B \; | \; \Value \; | \; \Any & \textit{key types}\\
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
S ::= & \;\; T &\\
& | \; T* & \textit{vararg types}\\
& | \; T \times S & \textit{tuple types}\\
P ::= & \Void \; | \; S & \textit{parameters list type}\\
R ::= & \Void \; | \; S \; | \; S \cup R & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

\section{Subtyping}

@@

\senv \vdash {\it number} \subtype \Number

@@

\senv \vdash {\it string} \subtype \String

@@

\section{Consistent-Subtyping}

@@

\senv \vdash {\it number} \lesssim \Number

@@

\senv \vdash {\it string} \lesssim \String

@@

\section{Typing rules}

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = \vec{e}  \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n}{:}\vec{T} = \vec{e} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}T = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; \vec{e} \; | \;
a\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it number} \; | \;
{\it string} \; | \;
{...} \; | \;
f \; | \;
\{ \; \vec{c}\; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
a \; | \;
l \; | \;
(e) \; | \;
{<}t{>} \; e\\
l ::= & \; n \; | \;
e_{1}[e_{2}]\\
a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
e_{1}{:}n(\vec{e_{2}})\\
f ::= & \; \mathbf{fun} \; (){:}R \; s \; | \;
\mathbf{fun} \; ({...}{:}T){:}R \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{T}){:}R \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{T},{...}{:}T){:}R \; s\\
c ::= & \; [e_{1}] = e_{2}\\
n ::= & \; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

### 2014-08-20 doc/thesis/typedlua.tex
Cosmetic changes to chapters 4 and 5

An important feature of Typed Lua is its backwards compatibility with Lua,
that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua.

@@

\begin{figure}[!ht]
\textbf{Type annotations}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Number} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}\\
\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()}, first-level types,
vararg types, tuple types, and union of tuple types.

@@

To match the semantics of Lua function calls, the default operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number, value*) -> (number, nil*)},
and the type of \texttt{distance} is \texttt{(any, any, value*) -> (number, nil*)}.
If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{absolute} receive any number of
extra arguments.
If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
replace the missing arguments and in the case of \texttt{absolute} generates
an error, as \texttt{nil} is neither a subtype of \texttt{number} nor consistent
with \texttt{number}.

Typed Lua also provides a strict operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number) -> (number)},
and the type of \texttt{distance} is \texttt{(any, any) -> (number)}.
If we call \texttt{absolute} with either extra or missing arguments,
Typed Lua generates errors about arity mismatch.

@@

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a type error
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom for giving a default value to an optional parameter. 
Typed Lua uses a typing rule to encode this idiom: 
if the left hand side of \texttt{or} has type \texttt{t|nil} and
the right hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv(dividend:number, divisor:number):
          (number, number)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) / divisor
        return q, r
      end 
    end
\end{verbatim}

@@

We could also have used \texttt{type(q) == "number"} or
even \texttt{type(r) == "number"} as the predicate of our example,
as both predicates would produce the same result.
However, the form that appears in our example is much more succint and idiomatic.

@@

The overloading mechanims of Typed Lua have a limitation:
the return type cannot depend on the input type.
Next, we show an example:
\begin{verbatim}
    local function limitation (x:number|string, y:number|string)
      if type(x) == "number" then
        if type(y) == "number" then
          return x + y
        end
      else
        if type(y) == "string" then
          return x .. y
        end
      end
    end
\end{verbatim}

This example shows that we cannot write a function that is
guaranteed to return a number when we pass two numbers and
guaranteed to return a string when we pass two strings.
Even though we can use intersection types to express the type
of this function as
\texttt{(number, number, value*) -> (number, nil*) \string&
(string, string, value*) -> (string, nil*)},
and thefore solving this limitation,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we are still working on this problem.

@@

In this example, the type of both variables \texttt{x} and \texttt{y}
is \texttt{string|nil}.
An incovenience aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

@@

Typed Lua includes \emph{interfaces} as syntactic sugar to table types
that specify records, as writing table types can be unwidely when
records get bigger and types of record fields get more complicated.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

@@

If our record type has optinal fields, that is, fields that can be \texttt{nil},
we need to use an explicit type declaration when declaring a
variable of this record type, as we show in the following example:
\begin{verbatim}
    local interface Person
      firstname:string
      middlename:string?
      lastname:string
    end
    local user1:Person = { firstname = "Lewis",
                           middlename = "Allan" or nil,
                           lastname = "Reed" }
    local user2:Person = { lastname = "Reed",
                           firstname = "Lou" }
\end{verbatim}

@@

We need to explicitly annotate \texttt{next} being of type
\texttt{Element|nil} because in our type system \texttt{nil}
is not the bottom type, that is, \texttt{nil} is subtype of itself only.
Typed Lua has this behavior because in Lua \texttt{nil} represents
the absence of an useful value.
Thus, making \texttt{nil} the bottom type would severaly reduce the amount
of type safety in the system.

Now we can use \texttt{Element} to annotate a function that
inserts an element at the begining of a list:
\begin{verbatim}
    local function insert (e:Element?, v:number):Element 
      return { info = v, next = e }
    end
\end{verbatim}

@@

Lua does not set policies on how programmers should define modules,
but it provides mechanisms for organizing a program in modules.
To load a module, Lua first checks whether the module is already
loaded or not.
When the module is not loaded, Lua executes its soure file as a
function, and the value returned by this function is the module.
Thus, Lua modules can be any Lua value.
However, our survey from Chapter \ref{chap:statistics} confirms that
the convention among Lua programmers is to use tables for defining modules. 
In this case, the fields of the table are the functions and other
values that the module exports.

@@

Another way of defining modules in Lua is to declare only locals
in the source file and return a table constructor that includes
the members that the module should export at the end of the source file.
The following example shows how Typed Lua handles this case:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    local function deg (x:number)
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number)
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number)
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

@@

The way Typed Lua handle modules using table types is also
relevant for supporting object-oriented programming, as we discuss in
the next section.

@@

While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
to check method calls, binding any occurence of the type \texttt{self}
in the type of the method to the receiver.
Indexing a method and not immediately calling it with the correct
receiver is a compile-time error:
\begin{verbatim}
    Shape.move(Shape, 10, 10)  -- Shape.x = 10 and Shape.y = 10
    Shape:move(5, 20)          -- Shape.x = 5 and Shape.y = 20
    local p = Shape.move       -- compile-time error
\end{verbatim}

@@

In the same example, we also declare the interface \texttt{Shape}.
Notice that we used double arrow instead of single arrow inside the
definition of the interface \texttt{Shape}.
The double arrow is syntactic sugar for defining the type of methods,
as it includes an implicit first parameter of type \texttt{self}.
The interface definition allows us to export the type of the class \texttt{Shape},
which is the same type that is given by the variable \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      const "new":(self, number, number) -> (self),
      const "move":(self, number, number) -> () }
\end{verbatim}

In Typed Lua, exporting interfaces allow us to use \texttt{require}
for creating prototype objects that work as classes,
along with a shortcut to the table type of the class.
This allows us to use this shortcut on type annotatoins,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

@@

The Typed Lua compiler also handles description files.
The programmer can use description files for two purposes:
export a statically typed interface for an existing Lua
module, without coverting it to Typed Lua,
or export a statically typed interface for a Lua module
that is written in C.

@@

The Typed Lua compiler searches for description files when it cannot
not find the respective Typed Lua file that is the argument of
\texttt{require}, as we can also use description files to give and
export the type of modules.

@@

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
A remarkable feature of Tidal Lock is the way that it handles tables,
because it allows the programmer to incrementally build the type of
a table.
More precisely, Tidal Lock allows the programmer to create an empty
table to build its type according to the keys that the programmer
inserts in it, so the programmer does not need to build the table in
a single table constructor.
Typed Lua has a simpler form of this feature, with a different
formalization.
Like Typed Lua, Tidal Lock also has local type inference.
Unlike Typed Lua, Tidal Lock does not type the common object-oriented
idioms used by Lua programmers.

>>>>

An important feature of Typed Lua is its backwards compatibility with Lua,
that is, unannotated Typed Lua code is valid Lua code.
For this reason, the syntactic extensions introduced by Typed Lua
include new keywords that are not reserved words.
Appendix \ref{app:syntax} presents the complete syntax of Typed Lua
in extended BNF.

@@

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} & ::= \; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} & ::= \; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Number} \; | \;
  \textit{String}\\
\textit{basetype} & ::= \; \textbf{boolean} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} & ::= \; \textit{tupletype} \; \texttt{`->'} \; \textit{rettupletype}\\
\textit{methodtype} & ::= \; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettupletype}\\
\textit{tabletype} & ::= \; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tupletype} & ::= \; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} & ::= \; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettupletype} & ::= \; \textit{unionlist} \; [\texttt{`?'}]\\
\textit{unionlist} & ::= \; \textit{tupletype} \; | \;
  \textit{unionlist} \; \texttt{`|'} \; \textit{unionlist}\\
\textit{tabletypebody} & ::= \; \textit{type} \; | \; \textit{fieldtypelist}\\
\textit{fieldtypelist} & ::= \; [\textbf{const}] \; \textit{fieldtype} \; \{\texttt{`,'} \; [\textbf{const}] \; \textit{fieldtype}\}\\ 
\textit{fieldtype} & ::= \; \textit{keytype} \; \texttt{`:'} \; \textit{type}\\
\textit{keytype} & ::= \; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{value} \; | \;
  \textbf{any}\\
\textit{rettype} & ::= \; \textit{type} \; | \; \textit{rettupletype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

Figure \ref{fig:types} presents the concrete syntax of the
Typed Lua types in extended BNF.
We classify Typed Lua types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types consist of \emph{type} and represent Lua values,
while second-level types consist of either \emph{tupletype} or
\emph{rettype} and represent the type of expression lists,
multiple assignments, and function applications.
First-level types include literal types, base types, the type \texttt{nil},
the top type \texttt{value}, the dynamic type \texttt{any},
the self type \texttt{self}, named types, function types, table types,
and union types.
Second-level types include the type void \texttt{()},
vararg types, tuple types, and union of tuple types.

@@

To match the semantics of Lua function calls, the \emph{default} operation
mode of Typed Lua adds a variadic tail to the parts of a function
type whenever the programmer does not specify one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number, value*) -> (number, nil*)},
and the type of \texttt{distance} is \texttt{(any, any, value*) -> (number, nil*)}.
If we call \texttt{absolute} with extra arguments, Typed Lua silently ignores
them, as the type signature lets \texttt{absolute} receive any number of
extra arguments.
If we call \texttt{absolute} with fewer arguments, Typed Lua uses \texttt{nil} to
replace the missing arguments and in the case of \texttt{absolute} generates
an error, as \texttt{nil} is neither a subtype of \texttt{number} nor consistent
with \texttt{number}.

Typed Lua also provides a \emph{strict} operation mode, and programmers can
use this mode if they want to catch arity mismatch of function calls.
In the strict mode, Typed Lua does not add a variadic tail to the
parts of a function type unless the programmer explicitly declares one.
In the previous examples, the type of both \texttt{factorial} and
\texttt{absolute} is \texttt{(number) -> (number)},
and the type of \texttt{distance} is \texttt{(any, any) -> (number)}.
If we call \texttt{absolute} with either extra or missing arguments,
Typed Lua generates errors about arity mismatch.

@@

Although the parameter \texttt{greeting} is optional, and has type
\texttt{string|nil}, the concatenation does not generate a type error
because we used the short-circuiting \texttt{or} operator to declare a
new variable \texttt{greeting} that is guaranteed to have type \texttt{string}. 
In Lua, only the values \texttt{nil} and \texttt{false} represent a
false condition, so programmers often use the \texttt{or} operator as a
common idiom to assign a default value to an optional parameter. 
Typed Lua uses a typing rule to encode this idiom: 
if the left hand side of \texttt{or} has type \texttt{t|nil} and
the right hand side has type \texttt{t} then the \texttt{or} expression has
type \texttt{t}.

@@

Lua programmers also overload the return type of functions,
usually for signaling the occurrence of errors.
In this idiom, a function returns its normal set of return values
when it successfully finished its execution,
and usually returns \texttt{nil} as the first value plus an
error message or other data that describes the error
when something failed during its execution.
Next, we show an example:
\begin{verbatim}
    local function idiv (dividend:number, divisor:number):
          (number, number)|(nil, string)
      if divisor == 0 then
        return nil, "division by zero"
      else
        local r = dividend % divisor
        local q = (dividend - r) / divisor
        return q, r
      end 
    end
\end{verbatim}

@@

We could also have used \texttt{type(q) == "number"} or
even \texttt{type(r) == "number"} as the predicate of our example,
as both predicates would produce the same result.
However, the form that appears in our example is much more succinct and idiomatic.

@@

The overloading mechanism of Typed Lua have a limitation:
the return type cannot depend on the input type.
Next, we show an example:
\begin{verbatim}
    local function limitation (x:number|string, y:number|string)
      if type(x) == "number" then
        if type(y) == "number" then
          return x + y
        end
      else
        if type(y) == "string" then
          return x .. y
        end
      end
    end
\end{verbatim}

This example shows that we cannot write a function that is
guaranteed to return a number when we pass two numbers and
guaranteed to return a string when we pass two strings.
Even though we could use intersection types to express the type
of this function as
\texttt{(number, number) -> (number) \string&
(string, string) -> (string)},
and therefore solve this limitation,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.

@@

In this example, the type of both variables \texttt{x} and \texttt{y}
is \texttt{string|nil}.
An inconvenience aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

@@

Typed Lua includes \emph{interfaces} as syntactic sugar to table types
that specify records, as writing table types can be unwieldy when
records get bigger and types of record fields get more complicated.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

@@

If our record type has optional fields, that is, fields that can be \texttt{nil},
we need to use an explicit type declaration when declaring a
variable of this record type, as we show in the following example:
\begin{verbatim}
    local interface Person
      firstname:string
      middlename:string?
      lastname:string
    end
    local user1:Person = { firstname = "Lewis",
                           middlename = "Allan" or nil,
                           lastname = "Reed" }
    local user2:Person = { lastname = "Reed",
                           firstname = "Lou" }
\end{verbatim}

@@

We need to explicitly annotate \texttt{next} being of type
\texttt{Element|nil} because in our type system \texttt{nil}
is not the bottom type, that is, \texttt{nil} is subtype of itself only.
Typed Lua has this behavior because in Lua \texttt{nil} represents
the absence of an useful value.
Thus, making \texttt{nil} the bottom type would severally reduce the amount
of type safety in the system.

Now we can use \texttt{Element} to annotate a function that
inserts an element at the beginning of a list:
\begin{verbatim}
    local function insert (e:Element?, v:number):Element 
      return { info = v, next = e }
    end
\end{verbatim}

@@

Lua does not set policies on how programmers should define modules,
but it provides mechanisms for organizing a program in modules.
To load a module, Lua first checks whether the module is already
loaded or not.
When the module is not loaded, Lua executes its source file as a
function, and the value returned by this function is the module.
Thus, Lua modules can be any Lua value.
However, our survey from Chapter \ref{chap:statistics} confirms that
the convention among Lua programmers is to use tables for defining modules. 
In this case, the fields of the table are the functions and other
values that the module exports.

@@

Another way of defining modules in Lua is to declare only locals
in the source file, and return a table constructor that includes
the members that the module should export at the end of the source file.
The following example shows how Typed Lua handles this case:
\begin{verbatim}
    local RADIANS_PER_DEGREE = 3.14159 / 180.0
    local function deg (x:number)
      return x / RADIANS_PER_DEGREE
    end
    local function rad (x:number)
      return x * RADIANS_PER_DEGREE
    end
    local function pow (x:number, y:number)
      return x ^ y
    end
    return {
      deg = deg,
      rad = rad,
      pow = pow,
    }
\end{verbatim}

@@

The way Typed Lua handles modules using table types is also
relevant for supporting object-oriented programming, as we discuss in
the next section.

@@

While \texttt{:} is syntactic sugar in plain Lua, Typed Lua uses it
to check method calls, binding any occurrence of the type \texttt{self}
in the type of the method to the receiver.
Indexing a method and not immediately calling it with the correct
receiver is a compile-time error:
\begin{verbatim}
    Shape.move(Shape, 10, 10)  -- Shape.x = 10 and Shape.y = 10
    Shape:move(5, 20)          -- Shape.x = 5 and Shape.y = 20
    local p = Shape.move       -- compile-time error
\end{verbatim}

@@

In the same example, we also declare the interface \texttt{Shape}.
Notice that we use double arrow instead of single arrow inside the
definition of the interface \texttt{Shape}.
The double arrow is syntactic sugar for defining the type of methods,
as it includes an implicit first parameter of type \texttt{self}.
The interface definition allows us to export the type of the class \texttt{Shape},
which is the same type that is given by the variable \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      const "new":(self, number, number) -> (self),
      const "move":(self, number, number) -> () }
\end{verbatim}

In Typed Lua, this style of classes definition allow us to use \texttt{require}
for creating prototype objects that work as classes,
along with a shortcut to the table type of the class.
This allows us to use this shortcut on type annotations,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

@@

The Typed Lua compiler also handles description files.
The programmer can use description files for two purposes:
export a statically typed interface for an existing Lua
module, without converting it to Typed Lua,
or export a statically typed interface for a Lua module
that is written in C.

@@

Figure \ref{fig:tldsyntax} shows the complete syntax of Typed Lua
description files in extended BNF.
A Typed Lua description file defines the table type that represents
a certain module and the type names that are exported
along with this table type.
We can export a type name through the declaration of an interface or
userdata.
As we saw, an interface declaration is just a shortcut to a table type.
An userdata declaration is a shortcut to a table type that includes
a \emph{brand} that is its name.
The Typed Lua compiler uses this brand to combine structural with
nominal type checking, so two different userdata are not subtype
of each other because they do not share the same brand.
The Typed Lua compiler searches for a description file when it cannot
not find the respective Typed Lua file that is the argument of
\texttt{require}.

@@

Tidal Lock \citep{tidallock} is a prototype of another optional type
system for Lua, which is written in Metalua.
A remarkable feature of Tidal Lock is the way that it handles tables,
because it allows the programmer to incrementally build the type of
a table.
More precisely, Tidal Lock allows the programmer to create an empty
table to build its type according to the keys that the programmer
inserts in it, so the programmer does not need to build the table in
a single table constructor.
This feature inspired us to introduce it into Typed Lua,
but in a simpler way and with different formalization.
Like Typed Lua, Tidal Lock also has local type inference.
Unlike Typed Lua, Tidal Lock does not type the common object-oriented
idioms used by Lua programmers, though it also uses subtyping.
